---
title: "DataExploration 5 yrs"
author: "LFM"
date: "4/8/2019"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r import, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(dplyr)
library(tidyverse)
library(tidyr)
library(lubridate)
getwd()

Pb1981 <- read.csv("EPA_Pb_Detroit1981_raw.csv")
Pb1985 <- read.csv("EPA_Pb_Detroit1985_raw.csv")
Pb1990 <- read.csv("EPA_Pb_Detroit1990_raw.csv")
Pb1995 <- read.csv("EPA_Pb_Detroit1995_raw.csv")
Pb2000 <- read.csv("EPA_Pb_Detroit2000_raw.csv")
Pb2005 <- read.csv("EPA_Pb_Detroit2005_raw.csv")
Pb2010 <- read.csv("EPA_Pb_Detroit2010_raw.csv")
Pb2015 <- read.csv("EPA_Pb_Detroit2015_raw.csv")
Pb2018 <- read.csv("EPA_Pb_Detroit2018_raw.csv")
```

```{r summary info}
str(Pb1981)
class(Pb1981$Date)
head(Pb1981)
summary(Pb1981)
```
#The dates shoud be Year/month/day
```{r attempt to Function Date format for all datasets}
#how to make this a function?
x <- Pb1981$Date

fmt <- as.Date(x, format = "%m/%d/%Y")

PbYears.Date <- function(x) {
    fmt <- as.Date(x, format = "%m/%d/%Y")
fmt
  }
PbYears.Date(c(Pb1981, Pb1985, Pb1990, Pb1995,
               Pb2000, Pb2005, Pb2010, Pb2015, 
               Pb2018))
```

```{r manual Date format}
Pb1981$Date <- as.Date(Pb1981$Date, format = "%m/%d/%Y")
Pb1985$Date <- as.Date(Pb1985$Date, format = "%m/%d/%Y")
Pb1990$Date <- as.Date(Pb1990$Date, format = "%m/%d/%Y")
Pb1995$Date <- as.Date(Pb1995$Date, format = "%m/%d/%Y")
Pb2000$Date <- as.Date(Pb2000$Date, format = "%m/%d/%Y")
Pb2005$Date <- as.Date(Pb2005$Date, format = "%m/%d/%Y")
Pb2010$Date <- as.Date(Pb2010$Date, format = "%m/%d/%Y")
Pb2015$Date <- as.Date(Pb2015$Date, format = "%m/%d/%Y")
Pb2018$Date <- as.Date(Pb2018$Date, format = "%m/%d/%Y")
```


```

#Gather data since there's some columns that are all the same information, or just not useful data (State, County, Daily Observation Count, Daily AQI, Percent Complete, etc.)
```{r}
Pb1981.gathered <- Pb1981 %>%
  select(Date, Site.ID, Site.Name, Daily.Mean.Pb.Concentration, UNITS,
         SITE_LATITUDE, SITE_LONGITUDE)
```
```{r}

```

#Explore the data graphically
```{r graphical representation}
library(colormap)

ggplot(Pb1981.gathered) +
  geom_histogram(aes(x = Daily.Mean.Pb.Concentration), bins = 45)

ggplot(Pb1981.gathered, aes(x = Date, y = Daily.Mean.Pb.Concentration)) +
  geom_col()

ggplot(Pb1981.gathered, aes(x = Date, y = Daily.Mean.Pb.Concentration)) +
  geom_point()

ggplot(Pb1981.gathered, aes(x = Date, y = Daily.Mean.Pb.Concentration), fill = Site.ID) +
  geom_point() +
  scale_y_log10()

ggplot(Pb2018, aes(x = Date, y = Daily.Mean.Pb.Concentration)) +
  geom_point(color = "Site.Name") +
  scale_fill_gradientn(colors = topo.colors(heat.colors))
```

## use rbind to put all 9 datasets together, then select columns

```{r compile into 1 dataset}
Pb1981to2018.gathered <- rbind(Pb1981, Pb1985, Pb1990, Pb1995, Pb2000, Pb2005,
                               Pb2010, Pb2010, Pb2015, Pb2018) %>%
   select(Date, Site.ID, Site.Name, Daily.Mean.Pb.Concentration, UNITS,
         SITE_LATITUDE, SITE_LONGITUDE)

write.csv()
```
2
## turn lat/long into points to map where in the city samples are from
# what are the highest concentrations of Pb in air? Are they near a certain spot?
# select what are highest concentrations (for example, highest 10% of samples)

```{r }

```

# should I change the UNITS to something else?

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
